Useful R tidbits: Social Networks

### 2022-03-20 _academics/ms797-glossary/ms797-glossary.Rmd
build

variance (34)
: _"the amount by which $\hat{f}$ would change if we estimated it using a different training data set"_

bias (35)
: _"the error that is introduced by approximating a real-life problem, which may be extremely complicated, by a much simpler model"_, as in the error from the (presumed) linearity of a regression against non-linear data whose complexity it does not capture. More flexible models increase variance and decrease bias.

@@

residual (61)
: the difference between $y_i$ and $\hat{y}_i$, also $e_i$; the difference between the $i$th response variable and the $i$th response variable predicted by the model

@@

ROC curve (150)
: Receiver Operating Characteristics; name of a curve showing the overall performance of a classifier resembling the top left corner of a rounded rectangle; area under the curve (AUC) indicates the percentage of correct classifications; the closer it is to square, the bigger the AUC, the better the classifier

>>>>

variance (34)
: _"the amount by which $\hat{f}$ would change if we estimated it using a different training data set"_
More practically: the *average* of squared differences from the mean, often expressed as $\sigma^2$, where $\sigma$ (or the square root of the variance) is the standard deviation
per [StatQuest](https://www.youtube.com/watch?v=EuBBz3bI-aA&t=246s): "the difference in fits between data sets" (like training and test)

bias (35)
: _"the error that is introduced by approximating a real-life problem, which may be extremely complicated, by a much simpler model"_, as in the error from the (presumed) linearity of a regression against non-linear data whose complexity it does not capture. More flexible models increase variance and decrease bias.
per [StatQuest](https://www.youtube.com/watch?v=EuBBz3bI-aA&t=136s): "The inability for a machine learning method (like linear regression) to capture the true relationship is called **bias**" - a straight line trying to model a curved separation in classes will never get it right and always be biased

@@

residual (61)
: the difference between $y_i$ and $\hat{y}_i$, also $e_i$; the difference between the $i$th response variable and the $i$th response variable predicted by the model
[StatQuest](https://youtu.be/g9c66TUylZ4?t=542): the difference between the *observed* and *predicted* value

@@

Note: sensitivity and specificity can and should be separately considered when evaluating ML methods. Each column (class) of a confusion matrix has a sensitivity and a specificity. In a 2x2, this is simple; in larger than 2x2, it involves summation.

$$\text{Sensitivity} = \frac{\text{True Positives}}{\text{True Positives + False Negatives}}$$
True positives is the cell in the class column containing the number of correct predictions in a category; the false negatives are the rest of the column.

$$\text{Specificity} = \frac{\text{True Negatives}}{\text{True Negatives + False Positives}}$$
True negatives is the sum of cells, in the other columns, that correctly do *not* predict the class (even if they wrongly predict some other class as well); the false positives are the rest of the *row* for the class that incorrectly predicted that class.

ROC charts show the true positive rate (sensitivity) on the Y axis, and the false positive rate (1 - sensitivity) on the X axis.

ROC curve (150)
: Receiver Operating Characteristics; name of a curve showing the overall performance of a classifier resembling the top left corner of a rounded rectangle; area under the curve (AUC) indicates the percentage of correct classifications; the closer it is to square, the bigger the AUC, the better the classifier. The Y axis is true positive rate (sensitivity). the X axis is false positive rate (1 - sensitivity).

[StatQuest](https://www.youtube.com/watch?v=4jRBRDbJemM&list=PLblh5JKOoLUICTaGLRoHQDuF_7q2GfuJF&index=6): AUC (area under the curve) of the ROC shows the overall effectiveness / quality of a model; good models will fill the space more. And the points of the individual curve help indicate the best thresholds to pick for the classifier; points with X = 0 have no false positives, and points with Y = 1 get all of the true positives.

Also, *precision* is another option for the X-axis, rather than false positives, if it's more important to know the proportion of the true positives were correctly identified.

$$\text{Precision} = \frac{\text{True Positives}}{\text{True Positives + False Positives}}$$

### 2022-03-20 _academics/ps753-glossary/ps753-glossary.Rmd
build

two-mode data / bipartite data
: ties between different sets (or categories) of agents, such as ties between nations and international organizations; a network split into two parts, each of whose nodes only connect to nodes in the opposite part, not to nodes within its own part

>>>>

two-mode data / bipartite data
: ties between different sets (or categories) of agents, such as ties between nations and international organizations; a network split into two parts, each of whose nodes only connect to nodes in the opposite part, not to nodes within its own part; "affiliation network" in @wassermanSocialNetworkAnalysis1994

### 2022-03-20 _academics/ps753-hw4/ps753-hw4.Rmd
build

This offers a very different view of the power dynamics in Florence. The Medici are greatly reduced, because their direct connections are all powerful families themselves, and so although they are well-connected by other measures, they may not be in a position of power over those other families. More influential families according to Bonacich would include the Ridolfi and Tornabuoni families, who suddenly become dominant from this perspective. The Bonacich calculation is complex and includes notions of centrality and power, and is not purely a measure of adjacency; see how small the Albizzi's Bonacich's score is, even though one of its three connections is to the pendant Ginori family, which is nonetheless in a slightly stronger position according to Bonacich. A fuller investigation of this dynamic would require some careful research.

>>>>

This offers a very different view of the power dynamics in Florence. The Medici are greatly reduced, because their direct connections are all powerful families themselves, and so although they are well-connected by other measures, they may not be in a position of power over those other families. More influential families according to Bonacich would include the Ridolfi and Tornabuoni families, who suddenly become dominant from this perspective. The Bonacich calculation is complex and includes notions of centrality and power, and is not purely a measure of adjacency; see how small the Albizzi's Bonacich score is, even though one of its three connections is to the pendant Ginori family, which is nonetheless in a slightly stronger position according to Bonacich. A fuller investigation of this dynamic would require some careful research.

### 2022-03-20 _academics/ps753-useful-r-tidbits/ps753-useful-r-tidbits.Rmd
build

@@

@@

## Transitivity

@@

# retrieve the index of isolate nodes
# (nodes with component count of 1 from "components" above)
isolates(gotbook.stat)
# There is no direct command in igraph, but we can do this:
# Create a list of the degree of each node in the network
deg_counts <- igraph::degree(gotbook.ig, loops = FALSE)
# filter and count the nodes with 0 degrees (or any other quantity of interest)
length(deg_counts[deg_counts == 0])

# subset vertex.names attribute to get names of isolates
as.vector(gotbook.stat %v% 'vertex.names')[c(isolates(gotbook.stat))]
```

@@

Again, there appear to be some inconsistency between `igraph` and `statnet` in the calculations, with statnet apparently not incorporating weights and failing on singluar matrices.

@@

>>>>

## Prelude: finding data to play with

Since in the class tutorials, the network objects are pre-loaded in the web-based runtime environments, we don't actually see how to load some of the data sets that the exercises are based on, for running outside of this environment.

There are gazillions of ways to get data into R, but for the purposes of playing with networks, there are some packages with pre-existing network data that can be easily loaded. Two of these are `igraphdata`, which contains igraph-formatted network datasets, and `ergm`, which contains statnet-compatible network datasets.

The datasets in each can be listed with the `data` command:

```r
> data(package = "igraphdata")

Data sets in package ‘igraphdata’:

Koenigsberg              Bridges of Koenigsberg from Euler's times
UKfaculty                Friendship network of a UK university faculty
USairports               US airport network, 2010 December
enron                    Enron Email Network
foodwebs                 A collection of food webs
immuno                   Immunoglobulin interaction network
karate                   Zachary's karate club network
kite                     Krackhardt's kite
macaque                  Visuotactile brain areas and connections
rfid                     Hospital encounter network data
yeast                    Yeast protein interaction network

> data(package = "ergm")

Data sets in package ‘ergm’:

cohab_MixMat (cohab)     Target statistics and model fit to a hypothetical
                         50,000-node network population with 50,000 nodes
                         based on egocent
cohab_PopWts (cohab)     Target statistics and model fit to a hypothetical
                         50,000-node network population with 50,000 nodes
                         based on egocent
cohab_TargetStats (cohab)
                         Target statistics and model fit to a hypothetical
                         50,000-node network population with 50,000 nodes
                         based on egocent
ecoli1 (ecoli)           Two versions of an E. Coli network dataset
ecoli2 (ecoli)           Two versions of an E. Coli network dataset
faux.desert.high         Faux desert High School as a network object
faux.dixon.high          Faux dixon High School as a network object
faux.magnolia.high       Goodreau's Faux Magnolia High School as a network
                         object
faux.mesa.high           Goodreau's Faux Mesa High School as a network
                         object
flobusiness (florentine)
                         Florentine Family Marriage and Business Ties Data
                         as a "network" object
flomarriage (florentine)
                         Florentine Family Marriage and Business Ties Data
                         as a "network" object
g4                       Goodreau's four node network as a "network"
                         object
kapferer                 Kapferer's tailor shop data
kapferer2 (kapferer)     Kapferer's tailor shop data
molecule                 Synthetic network with 20 nodes and 28 edges
samplike (sampson)       Cumulative network of positive affection within a
                         monastery as a "network" object
samplk1 (samplk)         Longitudinal networks of positive affection
                         within a monastery as a "network" object
samplk2 (samplk)         Longitudinal networks of positive affection
                         within a monastery as a "network" object
samplk3 (samplk)         Longitudinal networks of positive affection
                         within a monastery as a "network" object
```

To load the statnet "florentine" data from the `ergm` package:

```r
> data("florentine", package = "ergm")
```

This creates the objects `flobusiness` and `flomarriage`, both from this dataset, ready to go in statnet. Note that another version of this dataset is found in the `network` package `flo`:

```r
data("flo", package = "network")
```

This loads the very simple data into a matrix, but it is not yet a network object. To convert it to an igraph object:

```r
library(network)
library(igraph)

data(flo)
ntwk.ig <- graph_from_adjacency_matrix(flo)
```

@@

### Getting the basic network descriptors

First look at any network is to examine the network *size*, *type* (un/directed, un/weighted, bipartite) and available *attributes of vertices and edges*.

#### igraph

Basic descriptors first:

```r
vcount(climpref.ig)
[1] 34

ecount(climpref.ig)
[1] 531

is_bipartite(climpref.ig)
[1] FALSE

is_directed(climpref.ig)
[1] FALSE

is_weighted(climpref.ig)
[1] TRUE
```

Listing vertex and edge attributes:

```r
> vertex_attr_names(climpref.ig)
[1] "name"                 "Climate.council"      "Klimaallianz"        
[4] "Stiftung.Klimarappen" "type3"                "type5"               
[7] "dm"

> edge_attr_names(climpref.ig)
[1] "weight"

Accessing vertex and edge attributes:

```r
> V(climpref.ig)$name %>% head()
[1] "AA" "AB" "AC" "AD" "AE" "AF"

E(climpref.ig)$weight %>% head()
[1] 0.69 0.08 0.04 0.18 0.42 0.21
```

#### statnet

The basic descriptors in statnet are all *shown* by print():

```r
> print(flobusiness)
 Network attributes:
  vertices = 16 
  directed = FALSE 
  hyper = FALSE 
  loops = FALSE 
  multiple = FALSE 
  bipartite = FALSE 
  total edges= 15 
    missing edges= 0 
    non-missing edges= 15

 Vertex attribute names: 
    priorates totalties vertex.names wealth

No edge attributes
```

It appears, though it's thinly documented, that these attributes are programmatically accessible through the `$gal` attribute, as in `is_directed <- flobusiness$gal$directed`:

```r
> flobusiness$gal
$n
[1] 16

$mnext
[1] 16

$directed
[1] FALSE

$hyper
[1] FALSE

$loops
[1] FALSE

$multiple
[1] FALSE

$bipartite
[1] FALSE
```

@@

### Transitivity

@@

# retrieve the index of isolate nodes
# (nodes with component count of 1 from "components" above)
isolates(gotbook.stat)

# There is no direct command in igraph, but we can do this:
# Create a list of the degree of each node in the network
deg_counts <- igraph::degree(gotbook.ig, loops = FALSE)

# filter and count the nodes with 0 degrees (or any other quantity of interest)
length(deg_counts[deg_counts == 0])

# subset vertex.names attribute to get names of isolates
as.vector(gotbook.stat %v% 'vertex.names')[c(isolates(gotbook.stat))] %>%
  head()
##   [1] "Aegon Frey (Jinglebell)"         "Alebelly"                       
##   [3] "Alfyn"                           "Allar Deem"                     
##   [5] "Antario Jast"                    "Balman Byrch"    
```

@@

Again, there appear to be some inconsistency between `igraph` and `statnet` in the calculations, with statnet apparently not incorporating weights and failing on singular matrices.

@@

## Week 5: Big Block of Basic Code

Big Blocks of Basic Code to get a bunch of measures of a network:

```r
# Get the basic stuff we can do all at once with igraph
climinfl.nodes <- data.frame(
    name      = V(climinfl.ig)$name,
    totdegree = igraph::degree(climinfl.ig, loops=FALSE),
    indegree  = igraph::degree(climinfl.ig, mode="in", loops=FALSE),
    outdegree = igraph::degree(climinfl.ig, mode="out", loops=FALSE),
    eigen     = centr_eigen(climinfl.ig, directed = T)$vector,
    bonanich  = power_centrality(climinfl.ig),
    centr_clo = igraph::closeness(climinfl.ig),
    centr_btw = igraph::betweenness(climinfl.ig, directed = FALSE),
    # igraph only
    burt      = constraint(climinfl.ig)
)
# Network-level measures:
#   closeness centralization
climinfl.centr_clo = centr_clo(climpref.ig)$centralization
#   betweenness centralization
climinfl.centr_btw = centr_betw(climpref.ig, directed = FALSE)$centralization

# statnet version
climinfl.nodes <- data.frame(
    name      = climinfl.stat %v% "vertex.names",
    totdegree = sna::degree(climinfl.stat),
    indegree  = sna::degree(climinfl.stat, cmode = "indegree"),
    outdegree = sna::degree(climinfl.stat, cmode = "outdegree"),
    eigen     = evcent(climinfl.stat, ignore.eval = TRUE),
    bonanich  = bonpow(climinfl.stat),
    centr_clo = sna::closeness(climinfl.stat, gmode = "graph"),
    centr_btw = sna::betweenness(climinfl.stat, gmode = "graph")
)
# Network-level measures:
#   closeness centralization
climinfl.centr_clo = centralization(climinfl.stat, FUN = "closeness", mode = "graph")
#   betweenness centralization
climinfl.centr_btw = centralization(climinfl.stat, FUN = "betweenness", mode = "graph")

# Statnet-only: Gould-Fernandez Brokerage
temp <- data.frame(brokerage(climinfl.stat, cl = climinfl.nodes$orgtype5)$z.nli)
climinfl.nodes <- climinfl.nodes %>%
  mutate(broker.tot = temp$t,
         broker.coord = temp$w_I,
         broker.itin = temp$w_O,
         broker.rep = temp$b_IO,
         broker.gate = temp$b_OI,
         broker.lia = temp$b_O)

# Calculated measures not specific to igraph or statnet:
# Build the derived and reflected centrality (dc/rc) measures
# "To calculate the proportion of centrality that is received, we first
# square the adjacency matrix. The diagonal of the adjacency matrix is
# equal to the the square of node degree. We then divide this diagonal
# (squared degree) by total squared indegree (calculated by rowSums) to get
# the proportion of received centrality."

mat_climinfl <- as.matrix(as_adjacency_matrix(climinfl.ig))  # not " attr='weight'"
mat_climinfl_sq <- t(mat_climinfl) %*% mat_climinfl
climinfl.nodes$rc <- diag(mat_climinfl_sq) / rowSums(mat_climinfl_sq)
# replace missing values with 0
climinfl.nodes$rc <- ifelse(is.nan(climinfl.nodes$rc), 0, climinfl.nodes$rc)
climinfl.nodes$dc <- 1 - climinfl.nodes$rc

# Build the derived and reflected eigenvector measures
climinfl.nodes$eigen.rc <- climinfl.nodes$eigen * climinfl.nodes$rc
climinfl.nodes$eigen.dc <- climinfl.nodes$eigen * climinfl.nodes$dc
```

### Closeness centrailty

From text: "The closeness centrality of a node is defined as the sum of the geodesic distances between that node and all other nodes in a network."

Both are called in the above blocks, with `sna::closeness()` or `igraph::closeness()` on the respective network object.

From igraph::closeness help: "Closeness centrality measures how many steps is required to access every other vertex from a given vertex."

igraph and statnet have very different implementations, with options that have to be carefully set.

#### igraph

igraph uses inverse closeness.

For directed networks, use `mode=("in", "out", "all", "total")`, describing the path type; `in` is paths *to* a vertex, `out` is paths *from* a vertex. Undirected networks ignore this parameter. It will use the "weight" edge attribute automatically if it's there, or can be overriden with something else.

#### statnet

Must specify `gmode` (type of graph) as `graph` (undirected) or `digraph` (directed, default). `cmode` (type of closeness centrality being measured) is one of: `directed`, `undirected` (both standard closeness), `suminvdir` (directed case) and `suminvundir` (undirected case), and `gil-schmidt` for that. The `suminv` options correspond to igraph's default inversion, though they're still calculated slightly differently, so they're generally preferred.

statnet ignores the edge weights by default; `ignore.eval = FALSE` to use them, according to the documentation, but the results appear not to use them.

`sna::closeness(climpref.stat, gmode="graph", cmode="suminvundir", ignore.eval=FALSE))`

### Closeness centralization

Closeness centralization is the network-level measure of the *closeness centrality* node-level measure.

```r
climinfl.centr_clo = centr_clo(climpref.ig)$centralization
climinfl.centr_clo = centralization(climinfl.stat, FUN = "closeness", mode = "graph")
```

### Betweenness centrality

Betweenness centrality is the node-level measure of the number of geodesics (shortest path between two nodes) on which a node sits. A high betweenness centrality measure means a node is on many shortest-paths, suggesting a measure of influence or power.

```r
igraph::betweenness(climpref.ig, directed = FALSE)
sna::betweenness(climpref.ig, gmode = "graph")
```

The igraph version `directed` argument, for whether or not direction should be considered, defaults to `TRUE`; it might be wondered why it doesn't read the directedness of the graph as a default, but oh well.

The statnet version would use `gmode` of `digraph` for a directed network, and `cmode` for a variant undirected form; see `?sna::betweenness` for more. Statnet appears to use weights by default; `weights = NA` disables weights in igraph.

### Betweenness centralization

The network-level measure of betweenness centralization represents [Freeman centralization](https://en.wikipedia.org/wiki/Centrality#Freeman_centralization), at least according to the statnet documentation.

This is a "a measure of how central its most central node is in relation to how central all the other nodes are".

```r
climinfl.centr_btw = centr_betw(climpref.ig, directed = FALSE)$centralization
climinfl.centr_btw = centralization(climpref.stat, FUN = "betweenness", mode = "graph")
```

### (Burt's) network constraint

igraph-only function measuring a node's connection redundancy from 0 (none) to 1. Uses weights.

```r
constraint(climinfl.ig)
```

### Gould-Fernandez Brokerage

(Statnet-only)

From `?brokerage()`: "Gould and Fernandez (following Marsden and others) describe brokerage as the role played by a social actor who mediates contact between two alters."

Requires a directed network with a vertex attribute used for grouping. Returns a structure with a lot of information is returned; tutorial refers mainly to `znli` containing the following roles:

|prefix|Role|Action|Path|
|---|---|---|---|
|w_I|Coordinator| mediates contact between two individuals from his or her own group. |A -> A -> A|
|w_O|Itinerant broker| mediates contact between two individuals from a single group to which he or she does not belong. |A -> B -> A|
|b_{OI}|Gatekeeper| mediates an incoming contact from an out-group member to an in-group member. |A -> B -> B|
|b_{IO}|Representative| mediates an outgoing contact from an in-group member to an out-group member. |A -> A -> B|
|b_O|Liaison| mediates contact between two individuals from different groups, neither of which is the group to which he or she belongs. |A -> B -> C|
|t|Total|Total (cumulative) brokerage role occupancy|(Any two)|

```r
brokerage(climinfl.stat, cl = climinfl.nodes$orgtype5)
```
